cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-19 paper-cardinalR.Rmd
Updated the sepllings. Updated the summary sentence in the application.

---
title: "cardinalR: Generating Interesting High-Dimensional Data Structures"
abstract: >
  Simulated high dimensional data is useful for testing, validating, and improving algorithms used in dimensionality reduction, supervised and unsupervised learning. High-dimensional data is characterized by multiple variables that are dependent or associated in some way, such as linear, nonlinear, clustering or anomalies. Here we provide new methods for generating a variety of high-dimensional structures using mathematical functions and statistical distributions organized into the R package cardinalR. Several example data sets are also provided. These will be useful for researchers to better understand how different analytical methods work and can be improved, with a special focus on nonlinear dimension reduction methods. This package enriches the existing toolset of benchmark datasets for evaluating algorithms.
draft: true
author:
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
output:
  rjtools::rjournal_article:
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options:
  chunk_output_type: inline
---

@@

The example data structures are best viewed using a tour [@As85]. These show the data as a sequence of low dimensional projections (typically 2D), providing a good sense of the shape in high dimensions. The interactive tour plots included in this paper are produced using the software langevitour [@paul2023].

The paper is organized as follows. In the next section, we introduce the implementation of the `cardinalR` package on GitHub, including a demonstration of the package's key functions. We illustrate how a clustering data structure affects the dimension reductions in the Application section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for the use of our data collection.

@@

```{r branch, eval=knitr::is_html_output(), fig.cap="Viewing the $4\\text{-}D$ `orgcurvybranches` data. The tour view shows how the linear branches appear from multiple viewing angles.", fig.alt="Interactive langevitour of the orglinearbranches dataset in $4\\text{-}D$, showing linear branches from multiple viewing angles.", fig.pos="!ht", layout="l-body", fig.alt="The figure shows a tour of the 4‑D orgcurvybranches dataset. Points are displayed in a continuously changing 2‑D projection that rotates through the four dimensions. Each axis of the 2‑D view represents a linear combination of the original 4‑D coordinates. Points appear grouped into multiple elongated clusters or branches that move, rotate, and sometimes overlap as the projection changes. Branches are distinguished by colour, with individual points visible throughout. The display maintains a roughly square aspect ratio as the view transitions through multiple angles, illustrating the dataset’s branching structure from different perspectives."}

@@

```{r branch-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from the $4\\text{-}D$ \\texttt{orgcurvybranches} data. Each shows a different projection, illustrating how the linear branches appear from multiple viewing angles. These views highlight the dataset’s underlying branching structure and demonstrate how projections reveal patterns that are otherwise hidden in higher dimensions.", fig.width=12, fig.height=4, fig.alt="The figure consists of three 2‑D scatter plots, each showing a different 2‑D projection from the 4‑D orgcurvybranches dataset. In each plot, the horizontal axis represents one coordinate of the selected 2‑D projection, and the vertical axis represents the other coordinate. Points are plotted as individual markers and appear grouped into multiple elongated clusters or branches, which curve or radiate through the plot. Each branch is distinguished by colour, with overlapping branches occurring when subspaces are shared. The plots have roughly square aspect ratios, and together they show how the same high‑dimensional branching structure appears differently from three viewing angles.", fig.pos="!ht"}

@@

```{r cone, eval=knitr::is_html_output(), fig.cap="Viewing the $4\\text{-}D$ output  of `cone` data. Points are concentrated near the tip along the height dimension, while the radius of the hyperspherical cross-section decreases linearly toward the apex.", fig.alt="The figure shows a tour of the 4‑D cone dataset. Points are displayed in a continuously changing 2‑D projection, where each axis represents a linear combination of the original four dimensions. Points are densely packed near one end, forming a narrow tip, and gradually spread out along the other direction, creating a wider base. All points share the same colour and marker. As the projection rotates, the cone-like structure shifts in orientation, maintaining roughly square aspect ratios throughout the animation and showing the distribution of points from multiple viewing angles.", fig.pos="!ht"}

@@

```{r cone-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from the $4\\text{-}D$ \\texttt{cone} data. Points are concentrated near the tip along the height dimension, while the radius of the hyperspherical cross-section decreases linearly toward the apex. These projections show how the conical geometry is preserved.", fig.alt="The figure consists of three 2‑D scatter plots, each showing a different 2‑D projection of the 4‑D cone dataset. In each plot, the horizontal and vertical axes represent two continuous numeric dimensions. Points are plotted individually and are concentrated near one end, forming a narrow tip, and gradually spread out along the other direction, creating a wider base. The points trace an elongated, triangular or cone-like shape in each projection. All points use the same colour and marker, with no additional visual encodings, and the plots have roughly square aspect ratios. Together, the three projections illustrate the distribution of points along the conical structure from different viewing angles.", fig.pos="!ht", fig.width=12, fig.height=4}

@@

The `gen_longlinear(n, p)` function generates a high-dimensional dataset representing a single noisy linear trajectory. Let $t_i = i - 1, \quad i = 1, \dots, n$, denote a common latent index shared across all dimensions. For each dimension $j = 1, \dots, p)$, independent scale and shift parameters are sampled as $a_j \sim U(-10, 10), \qquad b_j \sim U(-300, 300)$. Gaussian noise $\varepsilon_{ij} \sim N(0, (0.03n)^2)$ is added independently across observations and dimensions. The observed variables are then defined as $X_{ij} = a_j \bigl(t_i + b_j + \varepsilon_{ij}\bigr), \quad i = 1, \dots, n$.cThis construction yields a single elongated linear structure embedded in $p\text{-}D$, with each dimension exhibiting a different orientation, scale, and offset.

@@

A polynomial structure generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in  space. To extend these patterns into high-dimensional settings, additional noise dimensions can be added. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity.

@@

```{r pyr, eval=knitr::is_html_output(), fig.cap="Viewing the $4\\text{-}D$  `pyrrect`, `pyrtri`, `pyrstar`, and `pyrholes` data. The `pyrrrect` structure forms a dense rectangular base tapering to a narrow tip, while `pytri` shows a more triangular spread with sharper edges. `Pyrstar` extends into multiple pointed branches radiating from a common core, and `pyrholes` reveals hollow or open regions within an otherwise compact shape.", fig.alt="The figure shows an interactive tour of the 4‑D pyrrrect, pyrtri, pyrstar, and pyrholes datasets. Points are displayed in continuously changing 2‑D projections, with each axis representing a linear combination of the four original dimensions. Points are plotted individually as small, uniformly coloured markers with no additional visual encodings. In pyrrrect, points form a dense rectangular base tapering to a narrow tip; in pyrtri, points form a triangular distribution with sharper edges; in pyrstar, points extend into multiple pointed branches radiating from a central region; and in pyrholes, points cluster in a compact shape but leave hollow or open regions within it. The projections rotate smoothly, showing the 4‑D structures from multiple viewing angles and maintaining roughly square aspect ratios throughout.", fig.pos="!ht"}

@@

```{r pyr-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the \\texttt{pyrrect} (a1-a3), \\texttt{pyrtri} (b1-b3), \\texttt{pyrstar} (c1-c3), and \\texttt{pyrholes} (d1-d3) data. The \\texttt{pyrrrect} structure forms a dense rectangular base tapering to a narrow tip, while \\texttt{pytri} shows a more triangular spread with sharper edges. \\texttt{pyrstar} extends into multiple pointed branches radiating from a common core, and \\texttt{pyrholes} reveals hollow or open regions within an otherwise compact shape. These projections illustrate a range of pyramid-like geometries that vary in density and structure.", fig.width=12, fig.height=16, out.width="80%", fig.align='center', fig.alt="The figure shows three 2‑D scatter plots for each of four 4‑D datasets: pyrrrect, pyrtri, pyrstar, and pyrholes. In each plot, the horizontal and vertical axes represent continuous numeric coordinates of the selected 2‑D projection, and points are plotted individually as small, uniformly coloured markers with no additional visual encodings. In pyrrrect, points form a dense rectangular base tapering to a narrow tip; in pyrtri, points form a triangular distribution with sharper edges; in pyrstar, points extend into multiple pointed branches radiating from a central region; and in pyrholes, points cluster in a compact shape but leave hollow or open regions within it. The plots maintain roughly square aspect ratios and illustrate how these different pyramid-like structures appear when projected from four to two dimensions.", fig.pos="!ht"}

@@

Building on simple circular structures, the `gen_unifsphere(n, r)` function function extends the idea to three dimensions by generating $n$ observations approximately uniformly distributed on the surface of a sphere of radius $r$.. Each observation is computed from spherical coordinates, with $u \sim U(-1, 1)$ representing $\cos(\phi)$ and $\theta \sim U(0, 2\pi)$ the azimuthal angle. Cartesian coordinates are then defined as $$X_1 = r\sqrt{1 - u^2}\cos(\theta), \quad X_2 = r\sqrt{1 - u^2}\sin(\theta),\text{ and }X_3 = ru,$$ ensuring uniform distribution on the surface (not within) of the sphere.

@@

In addition, the `gen_gridedsphere(n)` function constructs a $p$-dimensional dataset consisting of approximately $n$ points arranged on the surface of the unit $(p-1)\text{-}D$ sphere embedded in $\mathbb{R}^p$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) d"} else { "Figure \\@ref(fig:sphere-proj) d"})`). Rather than sampling points uniformly, this function creates a deterministic grid in spherical coordinates, using $(p-1)$ angular variables: the first $(p-2)$ angles are taken from $[0, \pi]$, and the final angle from $[0, 2\pi]$. The number of grid points along each angular dimension is determined by decomposing $n$ into $(p-1)$ approximately equal integer factors via `gen_nproduct(n, p - 1)`.

@@

```{r sphere-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, \\texttt{circle} (a1-a3), \\texttt{curvycycle} (b1-b3), and, $3\\text{-}D$ \\texttt{clusteredspheres} (c1-c3). The \\texttt{circle} structure forms a smooth, closed loop, while \\texttt{curvycycle} shows a wavy, continuous pattern forming a twisted ring. The \\texttt{clusteredspheres} dataset displays multiple compact spherical groups that are clearly separated in higher dimensions but overlap slightly in some $2\\text{-}D$ projections, highlighting how projection can distort spatial relationships. These projections show how simple cyclic, wavy curvilinear, and clustered structures appear in $2\\text{-}D$, emphasizing the effects of projection on density, continuity, and separation", fig.width=12, fig.height=12, out.width="80%", fig.align='center', fig.alt="A 3D scatterplot shows a projection of points sampled from a 4‑dimensional hemisphere onto three coordinate axes. Each point represents a location on the surface of a 4‑D unit sphere restricted to one hemisphere. The horizontal (x) and vertical (y) axes, along with depth (z), span roughly from –1 to 1, reflecting the Cartesian coordinates of the projected points. The dots are uniformly scattered over a curved, bowl-like region rather than filling a full sphere, with no strong clusters or gaps, illustrating that the sampling covers only one side of the sphere. If colour is used, it likely encodes an additional coordinate or density but does not introduce distinct categorical groups.", fig.pos="!ht"}

@@

Second, the `gen_curvycylinder(n, p, h)` function generates a $p$-dimensional dataset of $n$ observations structured as a $3\text{-}D$ cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when $p > 4$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) b"} else { "Figure \\@ref(fig:triginometric-proj) b"})`). The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension. Let $\theta \sim U(0, 3\pi)$ represent a random angle on a circular base and $z \sim U(0, h)$ represent the height along the cylinder. The coordinates are defined as: $X_1 = \cos(\theta)$ (Circular base, x-axis), $X_2 = \sin(\theta)$ (Circular base, y-axis), $X_3 = z$ (Linear height), and $X_4 = \sin(z)$ (Nonlinear curvy variation along height).

@@

Finally, the `gen_nonlinear(n, p, hc, non_fac)` function simulates a non-linear $2\text{-}D$ surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) f"} else { "Figure \\@ref(fig:triginometric-proj) f"})`). The $X_{1} \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors), $X_{3} \sim U(0.1, 0.8)$: independent auxiliary variable, $X_{2} = hc/X_{1} + \text{non\_fac}\sin(X_{1})$: non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and $X_{4} = \cos(\pi X_{1}) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter.

@@

High-dimensional data structures often benefit from the addition of auxiliary noise dimensions, which can be used to assess the robustness of dimensionality reduction and clustering algorithms. The functions in this section provide flexible ways to generate random noise dimensions, ranging from purely random Gaussian variables to more structured, wavy patterns that mimic non-linear distortions in high-dimensional space. These functions can be applied independently or combined with other geometric structures to create complex simulated datasets. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:noise-tb-html)', '\\@ref(tab:noise-tb-pdf)'))` details these functions.

@@

The `gen_wavydims2(n, p, x_1)` function extends this approach by applying a non-linear transformation to an existing dimension vector $x_1$:

@@

Finally, `gen_wavydims3(n, p, data)` generates noise for datasets with multiple correlated dimensions. The first three dimensions are small perturbations of the original coordinates $(X_1, X_2, X_3)$, while higher dimensions are constructed via non-linear combinations, including polynomial and trigonometric transformations, e.g.,

@@

```{r highd-data, eval=knitr::is_html_output(), fig.cap="Viewing five synthetic clusters with distinct geometric structures: a helical spiral, a hemisphere, a uniform cube, a cone, and a Gaussian cluster.", fig.alt="The figure shows a dynamic tour visualization of synthetic four-dimensional data displayed as a sequence of two-dimensional scatter plots. Each frame presents a different linear projection of the data, with the horizontal and vertical axes representing changing combinations of the original variables. Points represent individual observations and are coloured by cluster membership, with five distinct colours corresponding to the helical spiral, hemisphere, uniform cube, cone, and Gaussian clusters. As the tour progresses, the relative positions, shapes, and overlaps of the coloured point clouds change continuously, revealing elongated bands, curved arcs, compact clusters, and more diffuse regions depending on the projection. Clusters frequently move closer together or overlap in certain views, while becoming more separated in others, illustrating how apparent structure varies across different two-dimensional projections of the same high-dimensional data.", fig.pos="!ht"}

@@

As shown in Figure \@ref(fig:fig-nldr-layouts), tSNE (Figure \@ref(fig:fig-nldr-layouts) a) achieved the lowest HBE across bin widths (mostly tiny), indicating high preservation of both local and global structures. Its layout displays well-separated clusters with minimal inter-cluster distances, making it the most faithful representation of the underlying data structure. UMAP and PaCMAP (Figure \@ref(fig:fig-nldr-layouts) b and e) produced moderately accurate embeddings, although the six clusters appear more well-separated, while PHATE (Figure \@ref(fig:fig-nldr-layouts) c) show non-linear cluster structures irrespective of the original structure. Also, TriMAP (Figure \@ref(fig:fig-nldr-layouts) d) has high HBE, and show three clusters with small distances. PCA (Figure \@ref(fig:fig-nldr-layouts) f) failed to capture the non-linear geometry, leading to the highest HBE.

>>>>

---
title: "cardinalR: Generating Interesting High-Dimensional Data Structures"
abstract: >
  Simulated high dimensional data is useful for testing, validating, and improving algorithms used in dimension reduction, supervised and unsupervised learning. High-dimensional data is characterized by multiple variables that are dependent or associated in some way, such as linear, nonlinear, clustering or anomalies. Here we provide new methods for generating a variety of high-dimensional structures using mathematical functions and statistical distributions organized into the R package cardinalR. Several example data sets are also provided. These will be useful for researchers to better understand how different analytical methods work and can be improved, with a special focus on nonlinear dimension reduction methods. This package enriches the existing toolset of benchmark datasets for evaluating algorithms.
draft: true
author:
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
output:
  rjtools::rjournal_article:
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options:
  chunk_output_type: inline
---

@@

The example data structures are best viewed using a tour [@As85]. These show the data as a sequence of low dimensional projections (typically $2\text{-}D$), providing a good sense of the shape in high dimensions. The interactive tour plots included in this paper are produced using the software `langevitour` [@paul2023].

The next section provides an overview of the usage of the `cardinalR` package, illustrating how its modular components can be combined to generate complex high-dimensional datasets. This is followed by a section describing the implementation of the package, including its design principles and key functions. The Application section then demonstrates how the simulated clustering structures can be used to evaluate and compare dimension reduction and clustering methods. Finally, we give a brief conclusion of the paper and discuss potential opportunities for the use of our data collection.

@@

```{r branch, eval=knitr::is_html_output(), fig.cap="Viewing the $4\\text{-}D$ `orgcurvybranches` data. The tour view shows how the linear branches appear from multiple viewing angles.", fig.alt="Interactive langevitour of the orglinearbranches dataset in $4\\text{-}D$, showing linear branches from multiple viewing angles.", fig.pos="!ht", layout="l-body", fig.alt="The figure shows a tour of the 4‑D orgcurvybranches dataset. Points are displayed in a continuously changing 2‑D projection that rotates through the four dimensions. Each axis of the 2‑D view represents a linear combination of the original 4‑D coordinates. Points appear grouped into multiple elongated clusters or branches that move, rotate, and sometimes overlap as the projection changes. Branches are distinguished by color, with individual points visible throughout. The display maintains a roughly square aspect ratio as the view transitions through multiple angles, illustrating the dataset’s branching structure from different perspectives."}

@@

```{r branch-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from the $4\\text{-}D$ \\texttt{orgcurvybranches} data. Each shows a different projection, illustrating how the linear branches appear from multiple viewing angles. These views highlight the dataset’s underlying branching structure and demonstrate how projections reveal patterns that are otherwise hidden in higher dimensions.", fig.width=12, fig.height=4, fig.alt="The figure consists of three 2‑D scatter plots, each showing a different 2‑D projection from the 4‑D orgcurvybranches dataset. In each plot, the horizontal axis represents one coordinate of the selected 2‑D projection, and the vertical axis represents the other coordinate. Points are plotted as individual markers and appear grouped into multiple elongated clusters or branches, which curve or radiate through the plot. Each branch is distinguished by color, with overlapping branches occurring when subspaces are shared. The plots have roughly square aspect ratios, and together they show how the same high‑dimensional branching structure appears differently from three viewing angles.", fig.pos="!ht"}

@@

```{r cone, eval=knitr::is_html_output(), fig.cap="Viewing the $4\\text{-}D$ output  of `cone` data. Points are concentrated near the tip along the height dimension, while the radius of the hyperspherical cross-section decreases linearly toward the apex.", fig.alt="The figure shows a tour of the 4‑D cone dataset. Points are displayed in a continuously changing 2‑D projection, where each axis represents a linear combination of the original four dimensions. Points are densely packed near one end, forming a narrow tip, and gradually spread out along the other direction, creating a wider base. All points share the same color and marker. As the projection rotates, the cone-like structure shifts in orientation, maintaining roughly square aspect ratios throughout the animation and showing the distribution of points from multiple viewing angles.", fig.pos="!ht"}

@@

```{r cone-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from the $4\\text{-}D$ \\texttt{cone} data. Points are concentrated near the tip along the height dimension, while the radius of the hyperspherical cross-section decreases linearly toward the apex. These projections show how the conical geometry is preserved.", fig.alt="The figure consists of three 2‑D scatter plots, each showing a different 2‑D projection of the 4‑D cone dataset. In each plot, the horizontal and vertical axes represent two continuous numeric dimensions. Points are plotted individually and are concentrated near one end, forming a narrow tip, and gradually spread out along the other direction, creating a wider base. The points trace an elongated, triangular or cone-like shape in each projection. All points use the same color and marker, with no additional visual encodings, and the plots have roughly square aspect ratios. Together, the three projections illustrate the distribution of points along the conical structure from different viewing angles.", fig.pos="!ht", fig.width=12, fig.height=4}

@@

The `gen_longlinear(n, p)` function generates a high-dimensional dataset representing a single noisy linear trajectory. Let $t_i = i - 1, \quad i = 1, \dots, n$, denote a common latent index shared across all dimensions. For each dimension $j = 1, \dots, p)$, independent scale and shift parameters are sampled as $a_j \sim U(-10, 10), \qquad b_j \sim U(-300, 300)$. Gaussian noise $\varepsilon_{ij} \sim N(0, (0.03n)^2)$ is added independently across observations and dimensions. The observed variables are then defined as $X_{ij} = a_j \bigl(t_i + b_j + \varepsilon_{ij}\bigr), \quad i = 1, \dots, n$. This construction yields a single elongated linear structure embedded in $p\text{-}D$, with each dimension exhibiting a different orientation, scale, and offset.

@@

A polynomial structure generates data points that follow nonlinear curvilinear relationships, such as quadratic or cubic trends, in $2\text{-}D$ space. To extend these patterns into high-dimensional settings, additional noise dimensions can be added. These patterns are useful for evaluating how well algorithms capture smooth, nonlinear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity.

@@

```{r pyr, eval=knitr::is_html_output(), fig.cap="Viewing the $4\\text{-}D$  `pyrrect`, `pyrtri`, `pyrstar`, and `pyrholes` data. The `pyrrrect` structure forms a dense rectangular base tapering to a narrow tip, while `pytri` shows a more triangular spread with sharper edges. `Pyrstar` extends into multiple pointed branches radiating from a common core, and `pyrholes` reveals hollow or open regions within an otherwise compact shape.", fig.alt="The figure shows an interactive tour of the 4‑D pyrrrect, pyrtri, pyrstar, and pyrholes datasets. Points are displayed in continuously changing 2‑D projections, with each axis representing a linear combination of the four original dimensions. Points are plotted individually as small, uniformly colored markers with no additional visual encodings. In pyrrrect, points form a dense rectangular base tapering to a narrow tip; in pyrtri, points form a triangular distribution with sharper edges; in pyrstar, points extend into multiple pointed branches radiating from a central region; and in pyrholes, points cluster in a compact shape but leave hollow or open regions within it. The projections rotate smoothly, showing the 4‑D structures from multiple viewing angles and maintaining roughly square aspect ratios throughout.", fig.pos="!ht"}

@@

```{r pyr-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the \\texttt{pyrrect} (a1-a3), \\texttt{pyrtri} (b1-b3), \\texttt{pyrstar} (c1-c3), and \\texttt{pyrholes} (d1-d3) data. The \\texttt{pyrrrect} structure forms a dense rectangular base tapering to a narrow tip, while \\texttt{pytri} shows a more triangular spread with sharper edges. \\texttt{pyrstar} extends into multiple pointed branches radiating from a common core, and \\texttt{pyrholes} reveals hollow or open regions within an otherwise compact shape. These projections illustrate a range of pyramid-like geometries that vary in density and structure.", fig.width=12, fig.height=16, out.width="80%", fig.align='center', fig.alt="The figure shows three 2‑D scatter plots for each of four 4‑D datasets: pyrrrect, pyrtri, pyrstar, and pyrholes. In each plot, the horizontal and vertical axes represent continuous numeric coordinates of the selected 2‑D projection, and points are plotted individually as small, uniformly colored markers with no additional visual encodings. In pyrrrect, points form a dense rectangular base tapering to a narrow tip; in pyrtri, points form a triangular distribution with sharper edges; in pyrstar, points extend into multiple pointed branches radiating from a central region; and in pyrholes, points cluster in a compact shape but leave hollow or open regions within it. The plots maintain roughly square aspect ratios and illustrate how these different pyramid-like structures appear when projected from four to two dimensions.", fig.pos="!ht"}

@@

Building on simple circular structures, the `gen_unifsphere(n, r)` function extends the idea to three dimensions by generating $n$ observations approximately uniformly distributed on the surface of a sphere of radius $r$. Each observation is computed from spherical coordinates, with $u \sim U(-1, 1)$ representing $\cos(\phi)$ and $\theta \sim U(0, 2\pi)$ the azimuthal angle. Cartesian coordinates are then defined as $$X_1 = r\sqrt{1 - u^2}\cos(\theta), \quad X_2 = r\sqrt{1 - u^2}\sin(\theta),\text{ and }X_3 = ru,$$ ensuring uniform distribution on the surface (not within) of the sphere.

@@

In addition, the `gen_gridedsphere(n)` function constructs a $p$-dimensional dataset consisting of approximately $n$ points arranged on the surface of the unit $(p-1)$-sphere embedded in $\mathbb{R}^p$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) d"} else { "Figure \\@ref(fig:sphere-proj) d"})`). Rather than sampling points uniformly, this function creates a deterministic grid in spherical coordinates, using $(p-1)$ angular variables: the first $(p-2)$ angles are taken from $[0, \pi]$, and the final angle from $[0, 2\pi]$. The number of grid points along each angular dimension is determined by decomposing $n$ into $(p-1)$ approximately equal integer factors via `gen_nproduct(n, p - 1)`.

@@

```{r sphere-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, \\texttt{circle} (a1-a3), \\texttt{curvycycle} (b1-b3), and, $3\\text{-}D$ \\texttt{clusteredspheres} (c1-c3). The \\texttt{circle} structure forms a smooth, closed loop, while \\texttt{curvycycle} shows a wavy, continuous pattern forming a twisted ring. The \\texttt{clusteredspheres} dataset displays multiple compact spherical groups that are clearly separated in higher dimensions but overlap slightly in some $2\\text{-}D$ projections, highlighting how projection can distort spatial relationships. These projections show how simple cyclic, wavy curvilinear, and clustered structures appear in $2\\text{-}D$, emphasizing the effects of projection on density, continuity, and separation", fig.width=12, fig.height=12, out.width="80%", fig.align='center', fig.alt="A 3D scatterplot shows a projection of points sampled from a 4‑dimensional hemisphere onto three coordinate axes. Each point represents a location on the surface of a 4‑D unit sphere restricted to one hemisphere. The horizontal (x) and vertical (y) axes, along with depth (z), span roughly from –1 to 1, reflecting the Cartesian coordinates of the projected points. The dots are uniformly scattered over a curved, bowl-like region rather than filling a full sphere, with no strong clusters or gaps, illustrating that the sampling covers only one side of the sphere. If color is used, it likely encodes an additional coordinate or density but does not introduce distinct categorical groups.", fig.pos="!ht"}

@@

Second, the `gen_curvycylinder(n, p, h)` function generates a $p\text{-}D$ dataset of $n$ observations structured as a $3\text{-}D$ cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when $p > 4$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) b"} else { "Figure \\@ref(fig:triginometric-proj) b"})`). The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension. Let $\theta \sim U(0, 3\pi)$ represent a random angle on a circular base and $z \sim U(0, h)$ represent the height along the cylinder. The coordinates are defined as: $X_1 = \cos(\theta)$ (Circular base, x-axis), $X_2 = \sin(\theta)$ (Circular base, y-axis), $X_3 = z$ (Linear height), and $X_4 = \sin(z)$ (Nonlinear curvy variation along height).

@@

Finally, the `gen_nonlinear(n, p, hc, non_fac)` function simulates a nonlinear $2\text{-}D$ surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) f"} else { "Figure \\@ref(fig:triginometric-proj) f"})`). The $X_{1} \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors), $X_{3} \sim U(0.1, 0.8)$: independent auxiliary variable, $X_{2} = hc/X_{1} + \text{non\_fac}\sin(X_{1})$: nonlinear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and $X_{4} = \cos(\pi X_{1}) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a nonlinear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter.

@@

High-dimensional data structures often benefit from the addition of auxiliary noise dimensions, which can be used to assess the robustness of dimension reduction and clustering algorithms. The functions in this section provide flexible ways to generate random noise dimensions, ranging from purely random Gaussian variables to more structured, wavy patterns that mimic nonlinear distortions in high-dimensional space. These functions can be applied independently or combined with other geometric structures to create complex simulated datasets. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:noise-tb-html)', '\\@ref(tab:noise-tb-pdf)'))` details these functions.

@@

The `gen_wavydims2(n, p, x_1)` function extends this approach by applying a nonlinear transformation to an existing dimension vector $x_1$:

@@

Finally, `gen_wavydims3(n, p, data)` generates noise for datasets with multiple correlated dimensions. The first three dimensions are small perturbations of the original coordinates $(X_1, X_2, X_3)$, while higher dimensions are constructed via nonlinear combinations, including polynomial and trigonometric transformations, e.g.,

@@

```{r highd-data, eval=knitr::is_html_output(), fig.cap="Viewing five synthetic clusters with distinct geometric structures: a helical spiral, a hemisphere, a uniform cube, a cone, and a Gaussian cluster.", fig.alt="The figure shows a dynamic tour visualization of synthetic four-dimensional data displayed as a sequence of two-dimensional scatter plots. Each frame presents a different linear projection of the data, with the horizontal and vertical axes representing changing combinations of the original variables. Points represent individual observations and are colored by cluster membership, with five distinct colors corresponding to the helical spiral, hemisphere, uniform cube, cone, and Gaussian clusters. As the tour progresses, the relative positions, shapes, and overlaps of the colored point clouds change continuously, revealing elongated bands, curved arcs, compact clusters, and more diffuse regions depending on the projection. Clusters frequently move closer together or overlap in certain views, while becoming more separated in others, illustrating how apparent structure varies across different two-dimensional projections of the same high-dimensional data.", fig.pos="!ht"}

@@

As shown in Figure \@ref(fig:fig-nldr-layouts), tSNE (Figure \@ref(fig:fig-nldr-layouts) a) achieved the lowest HBE across bin widths (mostly tiny), indicating high preservation of both local and global structures. Its layout displays well-separated clusters with minimal inter-cluster distances, making it the most faithful representation of the underlying data structure. UMAP and PaCMAP (Figure \@ref(fig:fig-nldr-layouts) b and e) produced moderately accurate embeddings, although the six clusters appear more well-separated, while PHATE (Figure \@ref(fig:fig-nldr-layouts) c) show nonlinear cluster structures irrespective of the original structure. Also, TriMAP (Figure \@ref(fig:fig-nldr-layouts) d) has high HBE, and show three clusters with small distances. PCA (Figure \@ref(fig:fig-nldr-layouts) f) failed to capture the non-linear geometry, leading to the highest HBE.
